---
title: "Final Project - STAT 847"
subtitle: "Regression with an Abalone Dataset"
author: "Rawaha Nakhuda"
date: "2024-04-19"
output:
  github_document:
    fig_width: 8
    fig_height: 6
    dev: jpeg
---

\newpage

```{r, include = FALSE}
library(janitor)
library(compute.es) 
library(tidyverse)
library(car)
library(multcomp) 
library(pastecs)
library(tidyverse)
library(readxl)
library(broom)
library(skimr)
library(lubridate)
library(RColorBrewer)
library(gridExtra)
library(formattable)
library(knitr)
library(scales)
library(caret)
library(readxl)
library(factoextra)
library(GGally)
library(ggcorrplot)
knitr::opts_chunk$set(fig.height=6, fig.width=8, fig.align = "center", 
                      echo = TRUE, message = FALSE, warning  = FALSE, 
                      results = TRUE, dev = 'pdf')
select <- dplyr::select
opts <- options(knitr.kable.NA = "")
```

### Background

The [dataset](https://www.kaggle.com/competitions/playground-series-s4e4/) is part of the Kaggle Playground Series, Season 4 Episode 4. It describes physical measurements of Abalone, with the goal of the competition to predict its age (number of rings).

The dataset describes the following features:\
- **Sex**: Nominal — M, F, and I (infant)\
- **Length**: Continuous — mm — Longest shell measurement\
- **Diameter**: Continuous — mm — Measurement perpendicular to length\
- **Height**: Continuous — mm — Height with meat in shell\
- **Whole weight**: Continuous — grams — Whole abalone\
- **whole weight 1**: Continuous — grams — Weight of meat\
- **whole weight 2**: Continuous — grams — Gut weight (after bleeding)\
- **Shell weight**: Continuous — grams — Weight after being dried\
- **Rings**: Integer — +1.5 gives the age in years

```{r}
df <- read_csv("abalone.csv",
               show_col_types = FALSE) %>%
  clean_names() %>%
  select(-c(id))

df %>%
  skim_without_charts()
```

### 1. Approaches to the Dataset

**Approach 1:Prediction of Abalone Age**\
Using physical measurements to predict the age through the number of rings. This is the goal of the project, and would help marine scientists avoid the difficult process of cutting the abolone and counting the number of rings physically.

**Approach 2: Dimensionality Reduction for Prediction**\
Dimensionality reduction (e.g., PCA) can help reduce the number of predictor variables while retaining most of the information. This approach can help speed the performance of this model, and could group similar features (such as different weights) together.

\newpage
### 2. Cleaning Data

The data was imported using the Kaggle competition found [here](https://www.kaggle.com/competitions/playground-series-s4e4/). The data was relatively clean, and the following code was used to get it ready for basic analysis:


```{r, echo  = TRUE}
df <- read_csv("abalone.csv",
               show_col_types = FALSE) %>%
  clean_names() %>%
  select(-c(id))

# for basic data understanding:
df %>%
  skim_without_charts()
```

\newpage

### 3. Most Important Variables

To find the six most important variables, the first thing we will do is plot a correlation plot between all the variables with the number of rings (except ID). This will indicate the most important features for predicting age of Abolone.

```{r, results=TRUE, fig.cap = "Correlation Matrix of Abalone Dataset"}
cor_matrix <- (df) %>%
  select_if(is.numeric) %>%
  cor()

ggcorrplot(cor_matrix, hc.order = TRUE, type = "lower", lab = TRUE)
```

From the correlation plot, it seems the variables are highly correlated to each other. It seems the whole_weight variables are least correlated to the number of rings. 

The top six variables would then include the following:  
- sex, whole_weight, diameter, shell_weight, height and rings.  

These variables were removed for the following reason:  
- whole_weight_1 and whole_weight_2: high correlation to whole_weight and shell_weight and low correlation with rings.  
- length: 0.99 correlation to diameter. 

The ggpairs plot is then shown below:



```{r, fig.height=10, fig.width=10, results = TRUE, fig.cap="Pair Plot of 6 Most Important Variables"}
df %>%
  select(sex, diameter, height, whole_weight, shell_weight, rings) %>%
  mutate(sex = as.factor(sex)) %>%  # Ensure 'Sex' is treated as a factor
  ggpairs(mapping = ggplot2::aes(color = sex),
          upper = list(continuous = wrap("density", size = 1, alpha = 0.4),
                       combo = wrap("box_no_facet", size = 1, alpha = 0.4)),
          lower = list(continuous = wrap("points", size = 1, alpha = 0.4),
                       combo = wrap("dot_no_facet", size = 1, alpha = 0.4)),
          diag = list(
      continuous = wrap("densityDiag", alpha = 0.4),
      discrete = wrap("barDiag", alpha = 0.4)
    ))

abolone <- df %>%
  select(sex, diameter, height, whole_weight, shell_weight, rings)
```

The trends show that:  
- The 'I' sex shows significant deviation from the other M and F sex for mostly all variables.    
- The height column show the most number of outliers.  
- The rings column does not show significant linear patterns which matches the correlation plot previously, however the other variables do seem correlated with one other linearly.

\newpage

### 4. Classification Tree

In order to build a classification tree, we will split the rings column into a level of maturity, where greater than 10 rings will be classified as mature. Then we will develop our classification algorithm based on the other 5 variables. 

```{r, include = TRUE}
library(rpart)
library(caret)
library(rpart.plot)

df$mature <- ifelse(df$rings > 10, "mature", "immature")
df$mature <- as.factor(df$mature)


set.seed(11)  
train_index <- createDataPartition(df$mature, p = 0.80, list = FALSE)
training_data <- df[train_index, ]
testing_data <- df[-train_index, ]

# building tree
tree_model <- rpart(mature ~ sex + diameter + height + whole_weight + shell_weight, 
                    data = training_data, method = "class")
rpart.plot(tree_model, main="Classification Tree for Abalone Maturity", extra = 100)  


predictions <- predict(tree_model, testing_data, type = "class")
confusion_matrix <- confusionMatrix(predictions, testing_data$mature)

# predictions
example_data <- testing_data[1:2, ]
example_predictions <- predict(tree_model, example_data, type = "class")
```

```{r, results = TRUE}
#printing
kable(as.data.frame.matrix(confusion_matrix$table), caption = "Confusion Matrix for Abalone Maturity", align = 'c')
kable(data.frame(example_data[, c("sex", "diameter", "height", "whole_weight", "shell_weight", "mature")], Predicted = example_predictions), caption = "Prediction Table")
```

#### Following down the tree:

1. Shell Weight < 0.25 -> Immature  
2. Shell Weight < 0.25 -> Immature

\newpage

### 5. Continous Model

We will build a model to predict the rings column, from the other 5 most important variables. We will use PCA to reduce the dimensions of the dataset, and try to predict it using nonlinear terms as well!

```{r, echo = TRUE, results = TRUE}
ab_scaled <- abolone %>% 
  select(-c(sex, rings)) %>%
  scale()

pca_result  <- prcomp(ab_scaled, center = TRUE, scale. = TRUE)
summary(pca_result)
```
Since PC1 accounts for 94% of our variance, we can just use PC1 for our new model. 


```{r, echo = TRUE, results = TRUE}
abalone_pca <- data.frame(PC1 = pca_result$x[, 1], Rings = df$rings)
model <- lm(Rings ~ PC1 + I(PC1^2), data = abalone_pca)
kable(tidy(model), digits = c(1,2,5,1,40), caption = "Linear Regression Model Variables")
```

\newpage

### 6. 3 Variable ggplot

```{r, results = TRUE, fig.cap= "Relationship between Whole Weight, Rings, and Sex"}
ggplot(df, aes(x = whole_weight, y = rings, color = sex)) +
  geom_point(alpha = 0.5, size = 2) +  
  scale_color_manual(values = c("M" = "#00AFBB", "F" = "#E7B800", "I" = "#FC4E07")) +  
  labs(title = "Relationship between Whole Weight, Rings, and Sex",  
       x = "Whole Weight (grams)",
       y = "Rings",
       color = "Sex") +
  theme_minimal() +  
  theme(
    text = element_text(size = 14),  
    plot.title = element_text(hjust = 0.5),  
    panel.grid.minor = element_blank()  
  )
```


